%\documentclass[12pt]{aastex}
\documentclass[iop]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
\usepackage{paralist}
\shorttitle{Vertical shear instability}
\shortauthors{M-K. Lin, A. Youdin}

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12} 

\begin{document}

\title{Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula} 
\author{Min-Kai Lin, Andrew Youdin}
\affil{ $^1$Department of Astronomy and Steward Observatory, University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of accretion disks in which the disk  
  temperature $T$ is relaxed its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find that in a stably stratified disk,
  increasing $\beta$ even slightly away from    
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. 
  We apply our linear calculations  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI from $\sim 5$AU to $\sim 50$AU with  
  a characteristic growth time of $\sim 30$ local orbital 
  periods, and the dominant VSI mode in the MMSN is the `fundamental
  corrugation mode' which causes the disk column to oscillate
  vertically. We suggest the VSI is dynamically important in the outer
  parts of protoplanetary disks. 
\end{abstract}

\section{Introduction}
Understanding the mechanisms of angular momentum transport is the
central problem in the theory of accretion disks 
\citep{lyndenbell74}. The   
magneto-rotational instability in the context of differentially
rotating disks \citep[MRI,][]{balbus91} is now accepted as a robust
pathway to magneto-hydrodynamic (MHD) turbulence, which transports
angular momentum outwards and thus enables mass accretion. 

However, MHD turbulence may not be applicable to all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent  numerical simulations
suggest significant portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. Thus, alternative routes 
to angular momentum transport in PPDs remain of interest
\citep{armitage10}. 
 
One possibility is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height $z$ from the disk midplane. This
dependence occurs when the equilibrium pressure depends on both the
temperature and density, making the disk baroclinic. It is therefore
generally expected that $\p_z\Omega\neq 0$ in astrophysical disks.  

The VSI can be regarded as the disk analog of the
Goldreich-Schubert-Fricke instability 
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, can render
inertial waves unstable. \cite{barker15} gives a simple 
interpretation of the VSI: in an axisymmetric inviscid disk, the 
exchange of fluid parcels vertically, while conserving their specific angular
momentum, can result in a new configuration of lower energy if the
distribution of specific angular momentum varies with $z$. The VSI is
characterized by perturbations with large vertical length-scales
(relative to radial length-scales). 

Early analysis of the VSI in radially and vertically local 
models of accretion disks \citep{urpin98,urpin03} showed that the VSI
requires a short thermal timescale so that perturbations evolve 
isothermally, unless the disk is neutrally stratified in the vertical
direction. This is because vertical motions associated with the VSI
are stabilized by buoyancy forces if the disk is sub-adiabatically
stratified (convectively stable). A sufficiently short thermal
timescale renders buoyancy ineffective.    

The need for a short thermal timescale in stably stratified disks has
been demonstrated in numerical simulations of the VSI
\citep{nelson13,stoll14}. When the  thermal timescale $t_c$ is
parametrized in units of the orbital period  $P_\mathrm{orb}$,
\cite{nelson13} find that $t_c\lesssim  0.01P_\mathrm{orb}$ was needed
to observe the VSI. That is, the  disk temperature needs to remain
effectively unchanged. Radiation hydrodynamic simulations performed by
\cite{stoll14} also showed that the VSI is only sustained with the
inclusion of heating from stellar irradiation, which results in an
effective thermal timescale $t_c\ll P_\mathrm{orb}$ that maintains a
fixed temperature profile.     

Given that the VSI is generally sensitive to disk thermodynamics,  
it is important to understand and quantify this dependence, for
example on disk parameters, in order to evaluate the 
relevance of the VSI to realistic disks, in which one may expect
non-zero vertical entropy gradients and thermal timescales. Should a
given disk become turbulent through the VSI, it must first pass
through the linear instability. Thus, understanding the thermodynamic
dependence of the  linear VSI is essential.   

In this work we study the linear VSI in radially local, vertically
global models of astrophysical disks. This approach has
been adopted in recent studies of the linear VSI
\citep{nelson13,mcnally14,barker15}. However, these studies consider 
isothermal perturbations, for which the thermal timescale is formally
zero, and/or neutrally stratified disks in which the stabilizing
effect of bouyancy is absent. We account for both effects to study the
VSI in more general disk models. To do so, we examine 
sub-adiabatically stratified disks and include an energy equation 
with a source term that allows the thermodynamic response of 
perturbations to be varied smoothly between isothermal and 
adiabatic.   

% Our models are also an extension of earlier 
% analysis by \cite{urpin98} and \cite{urpin03}, which accounts for thermal
% evolution, but considered vertically localized disturbances.
% Our 
% vertically global approach is necessary to study disturbances with
% vertical wavelengths comparable to or larger than the disk scale
% height. Such disturbances were found by \cite{nelson13} and
% \cite{stoll14} to dominate their numerical simulations. We thus focus
% on vertically global perturbations which, in fact, we find to be the
% dominant mode in protoplanetary disks with a realistic estimate for
% the thermodynamic timescales. 

This paper is organized as follows. In \S\ref{setup} we list the
governing equations and global equilibrium disk models under
consideration. We formulate the linear stability problem in
\S\ref{linear} and simplify it by making the radially local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work and 
allows us to highlight the essential effects of vertical shear. In
particular, we quantify a characteristic upper limit to the thermal
timescale for the VSI to operate. In \S\ref{numerical} we numerically
solve the linear problem in the radially local approximation, but
relax all other approximations made in our analytical approach.    
We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale. In
\S\ref{application} we apply our linear models to the Minimum Mass
Solar Nebula (MMSN) by modeling thermal relaxation
timescales based on realistic disk properties, and find that the VSI may
operate in the outer parts of the MMSN. We summarize in
\S\ref{summary} with a discussion of caveats and possible directions
for future work.  

\input{setup}
\input{linear_iso}
\input{linear_adia}
\input{numerical}
\input{application} 
\input{summary}

\appendix
\input{appendix}

\bibliographystyle{apj}
\bibliography{ref}


\end{document}

